Proposal: A Natural Language Programming Toolkit for Python

### 2001-03-02 technical/proposal/proposal.tex
Added content:   * Wrote documentation section   * Wrote about half of the preliminary design & impl section
svn/trunk@9

\documentclass{article}
\usepackage{fullpage}

@@

The use of advanced language features and programming styles could
potentially make the toolkit much more difficult to use or to
understand.  The fewer language features that a student must learn in
order to use the toolkit, the faster they can start developing NLP
systems.  This section discusses the advantages and disadvantages of
using a variety of language features.  The language features used by
the toolkit are summarized by figure \ref{table:feature}.

\begin{figure}
\begin{centering}
\begin{tabular}{|l|l|}
\hline
\textbf{Language Feature} & \textbf{Use} \\
\hline
Classes & used \\
Exceptions & error conditions only \\
Operator Overloading & used \\
Python 2.0 Features & used \\
Function Arguments & minimal use \\
Lambda Functions & minimal use \\
Mapping and Filtering & not used \\
List Comprehensions & not used \\
\hline
\end{tabular}\\
\end{centering}
\caption{Summary of which advanced language features are used by the
toolkit.}
\label{table:feature}
\end{figure}

@@

Two main types of documentation:
  * tutorial docs
  * reference docs

reference docs are generated from inline comments by epydoc.

%################################################################
%#  PRELININARY DESIGN AND IMPLEMENTATION
%################################################################
\section{Preliminary Design and Implementation}

As part of the preliminary design process, I desinged and implemented
six prototype modules.

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{boxedminipage}

@@

The use of advanced language features and programming styles could
potentially make the toolkit much more difficult to use or to
understand.  The fewer language features that a student must learn in
order to use the toolkit, the faster they can start developing NLP
systems.  This section discusses the advantages and disadvantages of
using a variety of language features.  The language features used by
the toolkit are summarized by figure \ref{fig:feature}.

\begin{figure}
\begin{centering}
\begin{tabular}{|l|l|}
\hline
\textbf{Language Feature} & \textbf{Use} \\
\hline
Classes & used \\
Exceptions & error conditions only \\
Operator Overloading & used \\
Python 2.0 Features & used \\
Function Arguments & minimal use \\
Lambda Functions & minimal use \\
Mapping and Filtering & not used \\
List Comprehensions & not used \\
\hline
\end{tabular}\\
\end{centering}
\caption{Summary of which advanced language features are used by the
toolkit.}
\label{fig:feature}
\end{figure}

@@

The toolkit will be accompanied by documentation that explains the
toolkit, and describes how to use and extend it.  This documentation
will be divided into three primary categories:

\begin{itemize}
  \item \textbf{Technical Reports}, which explain and justify the
  toolkit's design and implementation.

  \item \textbf{Tutorials}, which teach students how to use the
  toolkit, in the context of performing specific tasks.

  \item \textbf{Reference Documentation}, which describes every
  module, interface, class, method, function, and variable in the
  toolkit.
\end{itemize}

Technical reports will be available as latex files, dvi files, and
postscript files.  Tutorials and reference documentation will be
available as HTML files, accessble over the web.

\subsection{Technical Reports}

Throughout the course of developing the toolkit, I will write a number
of technical reports, which are intended to explain and justify
specific aspects of the toolkit's design or implementation.  This
document is an example of a technical report.  Students should not
need to read the technical reports to use the toolkit.  However,
students can consult these reports if they would like further
information about how the tookit is designed and why it is designed
that way.

\subsection{Tutorials}

Tutorials are the primary type of documentation that students will use
to learn how to use the toolkit.  Each tutorial will explain how to
use some aspect of the toolkit, using specific examples.  In general,
the tutorials will explain how the student can perform some class of
tasks, such as using tokenizers and taggers to tag text with
part-of-speech tags.  Some tutorials may be associated with specific
toolkit modules.

Different tutorials will assume different amounts of background
knowledge about the toolkit.  This assumed background knowledge will
be specified by listing ``prerequisite'' tutorials for each tutorial.
At least one tutorial will assume no background knowledge about the
toolkit.

\subsection{Reference Documentation}

Reference documentation gives definitions for each module, interface,
class, method, function, and variable defined by the toolkit.  These
definitions should be both precise and consise.  I currently plan to
produce the reference documentation using Epydoc \cite{epydoc}.
Epydoc is a tool for automatically converting python documentation
strings to HTML pages.  Epydoc will generate the following HTML pages:

\begin{itemize}
  \item A page for each python \texttt{module}.  This page includes an
  overview of the \texttt{module}; a list of the \texttt{classs}es
  defined by the \texttt{module}; a list of the \texttt{function}s
  defined by the \texttt{module}, and a detailed description of each
  \texttt{function}; a list of the \texttt{variable}s defined by the
  \texttt{module}, and a detailed description of each
  \texttt{variable}.

  \item A page for each python \texttt{class}.  This page includes an
  overview of the \texttt{classs}; a list of the \texttt{method}s
  defined by the \texttt{classs}, and a detailed description of each
  \texttt{method}; a list of the instance \texttt{variable}s defined
  by the \texttt{classs}, and a detailed description of each instance
  \texttt{variable}; a list of the \texttt{classs} \texttt{variable}s
  defined by the \texttt{classs}, and a detailed description of each
  \texttt{classs} \texttt{variable}.

  \item A page containing the toolit's complete \texttt{classs}
  hierarchy.

  \item A page containing an index of every \texttt{module},
  \texttt{classs}, \texttt{method}, \texttt{function}, and
  \texttt{variable} defined by the toolkit.

\end{itemize}

Epydoc is currently under development.  Contact Edward Loper for more
information about Epydoc.  For examples of the HTML pages generated by 
Epydoc, see the reference documentation for the toolkit's prototype
modules:

 \texttt{http://www.cis.upenn.edu/$\sim$edloper/??}

%################################################################
%#  PRELIMINARY DESIGN AND IMPLEMENTATION
%################################################################
\section{Preliminary Design and Implementation}

As part of the preliminary design process, I desinged and implemented
six prototype modules.  These modules are summarized in figures
\ref{fig:mod-data} and \ref{fig:mod-task}, and described in the
following sections.  Most of these summaries are copied directly from
the modules' reference documentation.  For more information, see the
reference documentation at:

 \texttt{http://www.cis.upenn.edu/$\sim$edloper/??}

\begin{figure}
\noindent
\begin{centering}
\begin{boxedminipage}{.95\textwidth}
\begin{itemize}
  \item \textbf{Sets}: Encodes the mathmatical notion of a ``finite set.''
  \begin{itemize}
    \item \textit{Set}: A finite set.
  \end{itemize}

  \item \textbf{Tokens}: Encodes units of text such as words.
  \begin{itemize}
    \item \textit{TokenTypeI}:
         A unit of text.
    \item \textit{TextLocationI}:
         A location within a text.
    \item \textit{Token}:
         An occurance of a unit of text.
         Consists of a TokenType and a TextLocation.
  \end{itemize}

  \item \textbf{Syntax Trees}: Encodes syntax trees.  Not fully designed 
        yet.

  \item \textbf{Probability}: Encodes data structures associated with
        the mathmatical notion of probability, such as events and
        frequency distributions.
  \begin{itemize}
    \item Sample: Encodes the mathmatical notion of a
          ``sample.''  This is actually not implemented as a class or
          an interface -- (almost) anything can be a Sample.
    \item \textit{EventI}:
         A (possibly infinite) set of samples.
    \item \textit{FreqDistI}:
          The frequency distribution of a collection of samples.
    \item \textit{ProbDistI}:
          A probability distribution, typically derived from a
          frequency distribution (e.g., using ELE).
  \end{itemize}
\end{itemize}
\end{boxedminipage}\\
\end{centering}
 \caption[Data-oriented module prototypes]{Data-oriented module
 prototypes.  Under each module, the interfaces and top-level classes
 defined by that module are listed.}
\label{fig:mod-data}
\end{figure}

\begin{figure}
\noindent
\begin{centering}
\begin{boxedminipage}{.95\textwidth}
\begin{itemize}
  \item \textbf{Tokenizers}: Separate a string of text into a list of
       Tokens. 
  \begin{itemize}
     \item \textit{TokenizerI}
  \end{itemize}

  \item \textbf{Taggers}: Assign tags to each Token in a list of Tokens.
  \begin{itemize}
     \item \textit{TaggerI}
  \end{itemize}
\end{itemize}
\end{boxedminipage}\\
\end{centering}
 \caption[Task-oriented module prototypes]{Task-oriented module
 prototypes.  Under each module, the interfaces and top-level classes
 defined by that module are listed.}
\label{fig:mod-task}
\end{figure}

\subsection{Sets}

The set module contains a single class, \texttt{Set}.

\subsubsection{\texttt{Set} Class}

    An unordered container class that contains no duplicate elements.
    In particular, a set contains no elements $e_1$ and $e_2$ such
    that $e_1=e_2$.  Currently, the \texttt{Set} class is given a
    fairly minimal implementation.  However, more members may be
    defined in the future.

    Although the \texttt{Set} class attempts to ensure that it
    contains no duplicate elements, it can only do so under the
    following circumstances:

    \begin{itemize}
      \item For all elements $e_i$, $e_j$ added to the \texttt{Set},
           $e_i=e_j$ if and only if $e_j=e_i$.  This should always be the
           case as long as the elements in the \texttt{Set} use
           well-defined comparison functions.  An example where it
           would not be the case would be if ei defined
           \texttt{\_\_cmp\_\_()} to always return 0, and ej
           defined \texttt{\_\_cmp\_\_()} to always return $-1$.

      \item Mutable elements inserted in the \texttt{Set} are not
           modified after they are inserted.
    \end{itemize}

    If these circumstances are not met, the \texttt{Set} will
    continue to function, but it will no longer guarantee that it
    contains no duplicate elements.

\subsection{Tokens}

A Token is an occurance of a single unit of text, such as a word or a
punctuation mark.  The token module defines the \texttt{TokenTypeI}
interface, the \texttt{TextLocationI} interface, and the \texttt{Token
class}.

\subsubsection{\texttt{TokenTypeI} Interface}

    A \texttt{TokenTypeI} is single unit of text, such as a word or a
    punctuation mark.  The precise defintion of what counts as a token
    type will vary for different analyses.  For example, ``bank'' and
    ``run'' might be tokens for one analysis, while ``bank/N'' and
    ``bank/V'' might be tokens for another analysis.

    Token types can be compared for equality, and should be equal if
    and only if they refer to the same token type.

\vspace{2mm}\noindent
\textit{The \texttt{TokenTypeI} interface is currently implemented by
the classes \texttt{SimpleTokenType} and \texttt{TaggedTokenType}.}

\subsubsection{\texttt{TextLocationI} Interface}

    A \texttt{TextLocationI} is the location of an entity within a
    text.  Text locations can be compared for equality, and should be
    equal if and only if they refer to the same location.

\vspace{2mm}\noindent
\textit{The \texttt{TextLocationI} interface is currently implemented
by the \texttt{IndexTextLocation} class.}

\subsubsection{\texttt{Token} Class}

    A \texttt{Token} is an occurance of a single unit of text, such as
    a word or a punctuation mark.  A Token consists of a token type
    and a source.  The token type is the unit of text (e.g., a
    specific word).  The source is the position at which this token
    occured in the text.

    The precise defintion of what counts as a token type, or unit of
    text, will vary for different analyses.  For example, ``bank'' and 
    ``run'' might be tokens for one analysis, while ``bank/N'' and
    ``bank/V'' might be tokens for another analysis.

\subsection{Syntax Trees}

The syntax tree module defines data structures for storing syntax
trees.  It is still under development.  Currently, it defines one
interface: \texttt{SyntaxNodeI}

\subsubsection{\texttt{SyntaxNodeI} Interface}

\emph{This interface is still under development}

A syntax node is a single internal node of a syntax tree.  Each syntax
node is characterized by a tag and an ordered list of children.  The
tag specifies the type of the syntax Node (e.g., ``NP'' or ``V'').
Each child must either be a \texttt{Token}, or implement the
\texttt{SyntaxNodeI} interface.  Access to the children is implemented
by emulating a sequence type.  The \texttt{SyntaxNodeI} does
\emph{not} perform any semantic or syntactic checks on its arguments.
Figure \ref{fig:syntaxnode} shows how syntax nodes can be used:

\begin{figure}
\noindent
\begin{centering}
\begin{boxedminipage}{\textwidth}
\begin{tabbing}
\texttt{>>> cat = Token(\ldots)}\\
\texttt{>>> dog = Token(\ldots)}\\
\texttt{>>> the = Token(\ldots)}\\
\texttt{>>> chases = Token(\ldots)}\\
\\
\texttt{>>> cat\_np = SyntaxNode('NP', [the, cat])}\\
\texttt{>>> dog\_np = SyntaxNode('NP', [the, dog])}\\
\texttt{>>> vp = SyntaxTree('VP', [chases, dog\_np])}\\
\texttt{>>> s = SyntaxTree('S', [cat\_np, vp])}\\
\\
\texttt{>>> print s}\\
\texttt{\textit{[S [NP the dog] [VP chases [NP the dog]]]}}\\
\texttt{>>> print s[0]}\\
\texttt{\textit{[NP the dog]}}\\
\texttt{>>> print s[1].tag()}\\
\texttt{\textit{VP}}\\
\\
\texttt{>>> s[0] = dog\_np}\\
\texttt{>>> vp[:] = [cat, cat, cat]}\\
\end{tabbing}
\end{boxedminipage}\\
\end{centering}
 \caption[Example \texttt{SyntaxNode} use]
 {A session showing how \texttt{SyntaxNode}s can be used.}
\label{fig:syntaxnode}
\end{figure}

\subsection{Probability}

The probability module provides data structures associated with the
mathmatical notion of probability.  The following mathmatical terms
are useful for thinking about probability:

\begin{itemize}
  \item An \emph{experiment} is an occurance we observe whose result 
        is uncertain.
  \item An \emph{outcome} is some aspect of the result of an
        experiment which we wish to consider.
  \item A \emph{sample} is a possible outcome of an experiment.
  \item The \emph{sample space} of an experiment is the set of its
        possible outcomes.
  \item An \emph{event} is some subset of the sample space of an
        experiment. 
\end{itemize}

The toolkit does not currently encode the notion of a sample with an
interface; instead, it allows (almost) any object to function as a
sample.  Thus, for example, the sample space corresponding to a roll
of a die might be the set of integers from 1 to 6.  The only
requirement on samples is that they properly define the comparison
operator.

The toolkit currently defines the \texttt{CFSample} class for the
specific purpose of acting as a sample.  \texttt{CFSample}s consist of 
a (\emph{context}, \emph{feature}) pair.

The toolkit currently defines three interfaces for handling the
mathmatical notion of probability: the \texttt{EventI} interface; the
\texttt{FreqDistI} interface; and the \texttt{ProbDistI} interface.

\subsubsection{\texttt{EventI} Interface}

    An event is a subset of some sample space.  Note that this subset
    need not be finite.  Events are typically written as the set of
    samples they contain, or as a function in first order logic.
    Examples are:

\begin{verbatim}
{1,2,3}
{x:x>0}
\end{verbatim}

    The only method that events are required to implement is
    \texttt{\_\_contains\_\_()}, which tests whether a sample is a
    contained by the event.  However, when possible, events should
    also define the following methods:
    \begin{itemize}
      \item \texttt{\_\_cmp\_\_()}, which tests whether this event is
           equal to another event.
      \item \texttt{subset()}, which tests whether this event is a
           subset of another event.
      \item \texttt{superset()}, which tests whether this event is
           a superset of another event.
      \item \texttt{union()}, which returns an event containing the
           union of this event's samples and another event's samples.
      \item \texttt{intersection()}, which returns an event
           containing the intersection of this event's samples and
           another event's samples.
      \item \texttt{samples()}, which returns a \texttt{Set}
           containing all of the samples that are contained by this
           event. 
      \item \texttt{\_\_len\_\_()}, which returns the number of samples 
           contained by this event.
    \end{itemize}

    Classes implementing the \texttt{EventI} interface may choose
    to only support certain classes of samples, or may choose to only
    support certain types of events as arguments to the optional
    methods (\texttt{\_\_cmp\_\_}, \texttt{subset}, etc.).  If a
    method is unable to return a correct result because it is given an 
    unsupported type of sample or event, it should raise a
    NotImplementedError.

\vspace{2mm}\noindent
\textit{The \texttt{EventI} interface is currently implemented by the
classes: \texttt{UniversalEvent}, \texttt{NullEvent},
\texttt{SampleEvent}, \texttt{SetEvent}, \texttt{FuncEvent},
\texttt{ContextEvent}.}

\subsubsection{\texttt{FreqDistI} Interface}

(still writing documentation)

\subsubsection{\texttt{ProbDistI} Interface}

(still implementing)

\subsection{Tokenizers}

The process of separating a string of text into Tokens is called
\emph{tokenization}.  The tokenizer module defines the
\texttt{TokenizerI} interface, which should be implemented by classes
that tokenize text strings.

\subsubsection{\texttt{TokenizerI} Interface}

    A tokenizer is a processing class responsible for separating a
    string of text into a list of \texttt{Token}s.  This process
    is also known as \emph{tokenizing} the string of text.  Particular
    \texttt{Tokenizer}s may split the text at different points, or
    may produce Tokens with different types of \texttt{TokenType}.

\vspace{2mm}\noindent
\textit{The \texttt{TokenizerI} interface is currently implemented by
the classes \texttt{SimpleTokenizer} and \texttt{TaggedTokenizer}.}

\subsection{Taggers}

The process of assigning a tag to each Token in a

\subsubsection{\texttt{TaggerI} Interface}
